### 2014-04-03 ApTranscriptome_TR.Rmd
Updated analysis through comparison of thermal breadth for bimodally-expressed transcripts. Moved  function to scripts file.

interaction.transcripts <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.transcripts <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.transcripts))]
temperature.transcripts <- other.transcripts[which(Map(grepFunc, other.transcripts, term = "val") == TRUE)]
colony.transcripts <- other.transcripts[setdiff(names(other.transcripts), names(temperature.transcripts))]
rm(other.transcripts)
```

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for transcripts with interactions
interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.transcripts))

@@

pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")
```

@@

# Compare expression at optimum temp (A22.opt) between colonies using t-test
boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))

# T test
t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
```

@@

# Compare expression at optimum temp (Ar.opt) between colonies using t-test
boxplot(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))

# T test
t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))

@@

```{r intermediate_transcripts}
# list of transcripts that are 'Intermediate' expressed in Ar
Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
# T test
t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
wilcox.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)

# list of transcripts that are 'Intermediate' expressed in A22
A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
# T test
t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

@@

# calculate standard deviation of expression for each transcript
transcriptVar <- function(lmitem, colony) {
    # predict only for A22
    pred.vals <- seq(from = 0, to = 38.5, by = 0.5)
    newdf <- data.frame(val = pred.vals, colony = rep(colony, length = length(pred.vals)))
    pout <- predict(lmitem, newdata=newdf)
    pout <- cbind(newdf, pTPM = pout)

    # prediction can give values very close to zero or negative. biologically, these are meaningless so change to zero
    pout$pTPM <- ifelse(pout$pTPM < 0, 0, round(pout$pTPM, 5))

    # draw 1000 'temps' weighted by predicted expression
    random.draw <- sample(size = 1000, pred.vals, prob = pout$pTPM, replace = TRUE)

    # calculate standard deviation of weighted random.draw
    data.frame(exp_sd = sd(random.draw))
    } # end function

# apply function to all transcripts
A22.int.sd <- unlist(Map(transcriptVar, A22.int.lm, colony = "A22"))

# repeat for Ar
Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
Ar.int.sd <- unlist(Map(transcriptVar, Ar.int.lm, colony = "Ar"))

# T-test comparing standard deviation of expression between colonies
t.test(A22.int.sd, Ar.int.sd)

@@

Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.

~~Continue from here~~

@@

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 colony
A22trlist.bim <- A22.interaction.transcripts.type[which(A22.interaction.transcripts.type$A22.exp.type == "Bimodal"), "Transcript"]
A22.TPM.bim.dt <- TPM.dt.sub[A22trlist.bim]
setkey(A22.TPM.bim.dt, colony)
A22.TPM.bim.dt <- A22.TPM.bim.dt["A22"]
str(A22.TPM.bim.dt)

# calculate standard deviation of expression for each transcript
A22_bim_sd <- ddply(A22.TPM.bim.dt, .(Transcript), RxNsd.convex)

# repeat for Ar
Artrlist.bim <- Ar.interaction.transcripts.type[which(Ar.interaction.transcripts.type$A22.exp.type == "Bimodal"), "Transcript"]
Ar.TPM.bim.dt <- TPM.dt.sub[Artrlist.bim]
setkey(Ar.TPM.bim.dt, colony)
Ar.TPM.bim.dt <- Ar.TPM.bim.dt["Ar"]
str(Ar.TPM.bim.dt)

Ar_bim_sd <- ddply(Ar.TPM.bim.dt, .(Transcript), RxNsd.convex)
```

Next, I compare the standard deviations of expression for each transcipt between colonies.

```{r compare_Intermediate_variance}
# t-test
t.test(A22_bim_sd$exp_sd, Ar_bim_sd$exp_sd)

# Plot
# prep data
A22_bim_var$colony <- "A22"
Ar_bim_var$colony <- "Ar"
comb_bim_var <- rbind(A22_bim_var, Ar_bim_var)
# ggplot
g2 <- ggplot(comb_bim_var, aes(x=exp_sd, fill=colony)) + geom_density(alpha=0.2, position="identity")
g2 + scale_y_continuous(name="Density") +
  scale_x_continuous(name=expression("Standard deviation of expression function"))
```

Both the T-test and plot show that there is no difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.

>>>>

interaction.transcripts <- RxNlmAIC[which(Map(grepFunc, RxNlmAIC, term = "colonyAr:") == TRUE)]
other.transcripts <- RxNlmAIC[setdiff(names(RxNlmAIC), names(interaction.transcripts))]
temperature.transcripts <- other.transcripts[which(Map(grepFunc, other.transcripts, term = "val") == TRUE)]
colony.transcripts <- other.transcripts[setdiff(names(other.transcripts), names(temperature.transcripts))]
responsive.transcripts <- c(temperature.transcripts, interaction.transcripts)
rm(other.transcripts)
```

@@

```{r expression_shape, eval=TRUE, echo=TRUE}
# calculate response type for responsive transcripts
interaction.response.type <- ldply(interaction.transcripts, .progress = "text", .inform = TRUE, RxNtype)
stopifnot(nrow(interaction.response.type) == length(interaction.transcripts))

@@

pandoc.table(Ap.type.table, style="rmarkdown", caption = "Number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")

# only transcripts that differ in expression by colony
A22.interaction.type.table <- table(interaction.response.type[ , 'A22.type'])
Ar.interaction.type.table <- table(interaction.response.type[ , 'Ar.type'])

Ap.interaction.type.table <- rbind(A22.interaction.type.table, Ar.interaction.type.table)
rownames(Ap.interaction.type.table) <- c("A22", "Ar")

pandoc.table(Ap.interaction.type.table, style="rmarkdown", caption = "For transcripts that differ in expression pattern between *A22* and *Ar*, the number of transcripts with maximum expression at high, low, intermediate or both high and low (bimodal) temperatures.")

@@

# Compare expression at optimum temp (A22.opt) between colonies using t-test
t.test(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)
boxplot(A22.high.transcripts$A22.opt, A22.high.transcripts$Ar.opt)

# T test on log-transformed values to control for outliers
t.test(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
boxplot(log(A22.high.transcripts$A22.opt+1), log(A22.high.transcripts$Ar.opt+1))
```

@@

# compare expression at optimum temp (Ar.opt) between colonies using t-test
t.test(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
boxplot(Ar.low.transcripts$A22.opt, Ar.low.transcripts$Ar.opt)
# t-test with log values
t.test(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))
boxplot(log(Ar.low.transcripts$A22.opt+1), log(Ar.low.transcripts$Ar.opt+1))

@@

```{r intermediate_transcripts}
# list of transcripts that are 'Intermediate' expressed in Ar
Ar.int.transcripts <- Ap.response.type[which(Ap.response.type$Ar.type == "Intermediate"), ]
# T test
t.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt)
t.test(log(Ar.int.transcripts$A22.opt+1), log(Ar.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
wilcox.test(Ar.int.transcripts$A22.opt, Ar.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)

# list of transcripts that are 'Intermediate' expressed in A22
A22.int.transcripts <- Ap.response.type[which(Ap.response.type$A22.type == "Intermediate"), ]
# T test
t.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt)
t.test(log(A22.int.transcripts$A22.opt+1), log(A22.int.transcripts$Ar.opt+1))
# Wilcoxon signed rank-test
wilcox.test(A22.int.transcripts$A22.opt, A22.int.transcripts$Ar.opt, alternative = "two.sided", paired = TRUE, conf.int = TRUE)
```

@@

# apply `transcriptSD` function to all transcripts
A22.int.sd <- unlist(Map(transcriptSD, A22.int.lm, colony = "A22"))

# repeat for Ar
Ar.int.lm <- RxNlmAIC[Ar.int.transcripts$Transcript]
Ar.int.sd <- unlist(Map(transcriptSD, Ar.int.lm, colony = "Ar"))

# T-test comparing standard deviation of expression between colonies
t.test(A22.int.sd, Ar.int.sd)
```
Consistent with our hypothesis, 'Intermediate' transcripts in *Ar* are expressed over a significantly wider range of temperatures than in *A22*.

@@

@@

```{r calc_Bimodal_variance}
# extract 'Bimodal' expressed transcripts for A22 colony
A22.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$A22.type == "Bimodal"), "Transcript"]]
length(A22.bim.lm)

# apply `transcriptSD` function to all transcripts
A22.bim.sd <- unlist(Map(transcriptSD, A22.bim.lm, colony = "A22"))

# repeat for Ar 
Ar.bim.lm <- RxNlmAIC[Ap.response.type[which(Ap.response.type$Ar.type == "Bimodal"), "Transcript"]]
length(Ar.bim.lm)
Ar.bim.sd <- unlist(Map(transcriptSD, Ar.bim.lm, colony = "Ar"))

# t-test to compare standard deviation of 'Bimodal' transcripts between colonies
t.test(A22.bim.sd, Ar.bim.sd)
```

No difference in the standard deviation of expression for bimodally-expressed transcripts between colonies.
